\section{Conclusion} \label{sec:conc}
In this paper we present a model reduction routine that combines the classic model reduction method with respect to a weighted inner product with the symplectic model reduction. This allows the reduced system to be defined with respect to the norms and inner-products, natural to the problem. Furthermore, the symplectic nature of the reduced system will preserve the Hamiltonian structure of the original system, which result in robustness and enhanced stability in the reduced system.

We demonstrate that including the weighted inner-product in the syplectic model reduction can be viewed as a natural extension of the symplectic method. Therefore, the stability preserving properties of the symplectic method generalize naturally to the new method.

Numerical results suggest the classic model reduction methods with respect to an weighted inner product can help with the boundedness of the system. However, only the symplectic treatment can consistently increase the accuracy of the reduced system. This is consistent with the fact the symplectic methods preserve the Hamiltonian structure.

We also exhibit that to accelerate the evaluation of the nonlinear terms, adopting a symplastic approach is essential. This allows an accurate reduced system that is consistently enhanced when the basis for the nonlinear term is enriched.

Hence, the symplectic model-reduction with respect to a weigther inner product can provide an accurate and robust reduced system that carry the norms and inner products most appropriate to the problem.

\section*{Acknowledgments} We would like to show our sincere appreciation to Dr. Claudia Maria Colciago for the several brainstorming meetings which helped with the development of the main parts of this article. Furthermore, we would like to thank Prof. Bernard Haasdonk who initiated the investigation which led to this article. Finally, we would like to thank Prof. Karen Willcox for hosting Babak Maboudi Afkham at MIT during the composition of this paper. 
