%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%

%
\RequirePackage{fix-cm}

\makeatletter
\def\cl@chapter{}
\makeatother

%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
\usepackage{braket,amsfonts}

%% Used in table example below
\usepackage{array}

%% Used in table and figure examples below
\usepackage[caption=false]{subfig}
%% Used for papers with subtables created with the subfig package
%\captionsetup[subtable]{position=bottom}
%\captionsetup[table]{position=bottom}

%% Used for PgfPlots example, shown in the "Figures" section below.
\usepackage{pgfplots}

\usepackage{enumerate}

%% Algorithm style, could alternatively use algpseudocode
%\usepackage{algorithm}
%\usepackage{algpseudocode}

%% For figures
\usepackage{graphicx,epstopdf}

%% For creating math operators
\usepackage{amsopn}
\DeclareMathOperator{\Range}{Range}

% additional useful packages
\usepackage[disable]{todonotes}
%\usepackage{showlabels}
\usepackage{autonum}

\usepackage{cleveref}
\Crefname{table}{Algorithm}{Algorithms}

\usepackage{caption}

\makeatletter \let\cl@chapter\relax \makeatother
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
\newcommand{\edit}[1]{{\color{red} #1}}
\newcommand{\editA}[1]{{\color{blue} #1}}

\newcommand\bs{\symbol{'134}} % print backslash in typewriter OT1/T1
\newcommand{\preamble}[2][\small]{\textcolor{texpreamble}{#1\texttt{#2 \emph{\% <- Preamble}}}}

\spnewtheorem{rem}{Remark}[section]{\bf}{\it}
%\spnewtheorem{algorithm}{Algorithm}[section]{\bf}{\rm}
%
% Insert the name of "your journal" with
\journalname{Advances in Computational Mathematics}
%


\begin{document}

\title{Symplectic Model-Reduction with a Weighted Inner Product\thanks{Babak Maboudi Afkham is supported by the SNSF under the grant number P1ELP2\_175039. Ashish Bhatt and Bernard Haasdonk gratefully acknowledge the support of DFG grant number HA5821/5-1.}
}

%\titlerunning{Short form of title}        % if too long for running head

\author{Babak Maboudi Afkham \and
        Ashish Bhatt \and
        Bernard Haasdonk \and
        Jan S. Hesthaven
}

%\authorrunning{Short form of author list} % if too long for running head

\institute{B.M. Afkham, J.S. Hesthaven \at
              Institute of Mathematics (MATH), School of Basic Sciences (FSB), \\
              Ecole Polytechnique F\'ed\'erale de Lausanne, 1015 Lausanne, Switzerland \\
              \email{[babak.maboudi,jan.hesthaven]@epfl.ch}.
           \and
           A. Bhatt \at
              2nd floor Science Block, \\
			  IIT (ISM) Dhanbad, \\
              Dhanbad, India, 826004 \\
              \email{ashishbhatt@iitism.ac.in}
           \and
           B. Haasdonk \at
              Pfaffenwaldring 57,\\
              FB Math, University of Stuttgart,\\
              70569 Stuttgart, Germany \\
              \email{haasdonk@mathematik.uni-stuttgart.de}
}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor


\maketitle

\begin{abstract}
In the recent years, considerable attention has been paid to preserving 
structures and invariants in reduced basis methods, in order to enhance the 
stability and robustness of the reduced system. In the context of Hamiltonian 
systems, symplectic model reduction seeks to construct a reduced system that 
preserves the symplectic symmetry of Hamiltonian systems. However, symplectic 
methods are based on the standard Euclidean inner products and are not suitable 
for problems equipped with a more general inner product. In this paper we 
generalize symplectic model reduction to allow for the norms and inner products 
that are most appropriate to the problem while preserving the symplectic 
symmetry of the Hamiltonian systems. To construct a reduced basis and 
accelerate the evaluation of nonlinear terms, a greedy generation of a 
symplectic basis is proposed. Furthermore, it is shown that the greedy approach 
yields a norm bounded reduced basis. The accuracy and the stability of this 
model reduction technique is illustrated through the development of reduced 
models for a vibrating elastic beam and the sine-Gordon equation.
\keywords{Structure Preserving \and Weighted MOR \and Hamiltonian Systems \and Greedy Reduced Basis \and
Symplectic DEIM}
% \PACS{PACS code1 \and PACS code2 \and more}
\subclass{78M34 \and 34C20 \and 35B30 \and 37K05 \and 65P10 \and 37J25}
\end{abstract}

\input{./chapters/1.intro.tex}
\input{./chapters/2.hamil.tex}
\input{./chapters/3.mor.tex}
\input{./chapters/4.norm_mor.tex}
\input{./chapters/5.res.tex}
\input{./chapters/6.conc.tex}

%\begin{acknowledgements}
%If you'd like to thank anyone, place your comments here
%and remove the percent signs.
%\end{acknowledgements}

% BibTeX users please use one of
%\bibliographystyle{spbasic}      % basic style, author-year citations
\bibliographystyle{spmpsci}      % mathematics and physical sciences
%\bibliographystyle{spphys}       % APS-like style for physics
\bibliography{references}   % name your BibTeX data base

\end{document}
% end of file template.tex

